---
title: "Exploring Exercise Response Across Multi-Omics from a Custom Gene List"
author:
  - name: "David Jimenez-Morales"
    affiliation: "MoTrPAC Bioinformatics Center, Stanford University"
    email: "motrpac-helpdesk@lists.stanford.edu"
output:
  rmdformats::readthedown:
    toc_depth: 4
    code_folding: show
vignette: >
  %\VignetteIndexEntry{Exploring Exercise Response Across Multi-Omics from a Custom Gene List}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = 'center'
)
```

# Summary

This interactive workflow enables you to explore how **your genes of interest** respond to endurance exercise training across multiple tissues and molecular layers in the MoTrPAC rat study.

**What you'll get:**

- **Dataset Overview**: Visualization of all available multi-omics data (transcriptomics, proteomics, metabolomics, etc.)
- **Gene Coverage**: Automatic mapping of your genes across different omics platforms and tissues
- **Interactive Tables**: Searchable, filterable results showing log fold-changes and significance
- **Trajectory Plots**: Visualizations of how your genes change over the training time course (1w, 2w, 4w, 8w)
- **Heatmaps**: Sex-specific comparison of gene responses across tissues and time points with significance markers
- **Multi-Tissue Insights**: Identify tissue-specific and sex-specific exercise responses

**Just provide your gene list**, and this notebook will automatically retrieve and visualize endurance training effects across the comprehensive MoTrPAC dataset spanning 7 omics platforms and 20 tissues!

# Introduction

This vignette demonstrates how to explore a custom gene list across the MoTrPAC endurance exercise training multi-omics dataset. The workflow is designed to help you:

- **Input** your own gene list of interest
- **Retrieve** differential analysis results across multiple omics layers and tissues
- **Visualize** temporal trajectories showing how genes respond over 1-8 weeks of training
- **Compare** responses between males and females, and across different tissues
- **Identify** significant training-induced changes (FDR < 0.05)

## Required Packages

To facilitate access to and analysis of this extensive dataset, two
essential R packages are used:

-   [MotrpacRatTraining6moData](https://motrpac.github.io/MotrpacRatTraining6moData/):
    Provides access to the processed data and downstream analysis
    results from the MoTrPAC endurance exercise study.
-   [MotrpacRatTraining6mo](https://motrpac.github.io/MotrpacRatTraining6mo/):
    Offers functions to help retrieve and explore the data, enabling
    users to perform analyses and reproduce key findings from the study.

```{r setup, message=FALSE, warning=FALSE}
library(MotrpacRatTraining6moData)
library(MotrpacRatTraining6mo)

# additional R packages
library(ggplot2)
library(dplyr)
library(tidyr)
library(tibble)
library(pheatmap)
library(viridis)
library(DT)
library(gridExtra)
```

# Define Your Gene List

Start by defining a custom gene list. This can be any set of genes you're interested in exploring. Here we use an example list of genes related to metabolism and exercise response:

```{r define-gene-list}
# Example gene list - Replace with your genes of interest
custom_genes <- c(
  "Ppargc1a", 
  "Vegfa",
  "Hif1a",
  "Pdk4",
  "Ucp1",
  "Myh7",
  "Acadm", 
  "Slc2a4",
  "Cpt1b",
  "Nrf1"
)

cat("Your gene list contains", length(custom_genes), "genes:\n")
print(custom_genes)
```

# Retrieve Multi-Omics Data

## Overview of Available Datasets

Before diving into your custom gene list, let's explore what differential analysis datasets are available in the MoTrPAC rat endurance training study:

```{r dataset-overview, fig.width=12, fig.height=8}
# Load the differential analysis results
da_results <- combine_da_results()

# Get unique combinations of assay and tissue
dataset_summary <- da_results %>%
  select(assay, tissue) %>%
  distinct() %>%
  arrange(assay, tissue)

# Create summary counts
assay_counts <- dataset_summary %>%
  count(assay, name = "n_tissues") %>%
  arrange(desc(n_tissues))

tissue_counts <- dataset_summary %>%
  count(tissue, name = "n_assays") %>%
  arrange(desc(n_assays))

# Create tissue order based on number of assays
tissue_order <- tissue_counts %>%
  arrange(desc(n_assays)) %>%
  pull(tissue)

# Apply factor levels for proper ordering
dataset_summary_ordered <- dataset_summary %>%
  mutate(
    assay = factor(assay, levels = unique(assay)),
    tissue = factor(tissue, levels = tissue_order)
  )

# Get color mappings for available assays and tissues
assay_colors_used <- ASSAY_COLORS[names(ASSAY_COLORS) %in% unique(dataset_summary_ordered$assay)]
tissue_colors_used <- TISSUE_COLORS[names(TISSUE_COLORS) %in% unique(dataset_summary_ordered$tissue)]

# Create presence/absence heatmap with assay colors
phm <- ggplot(dataset_summary_ordered, aes(x = tissue, y = assay)) +
  geom_tile(aes(fill = assay), color = "white", linewidth = 1) +
  scale_fill_manual(values = assay_colors_used) +
  scale_y_discrete(limits = rev(levels(dataset_summary_ordered$assay))) +
  labs(
    title = "Available Differential Analysis Datasets",
    subtitle = "MoTrPAC Rat Endurance Training 6mo Data",
    x = "Tissue",
    y = "Assay"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5, color = "gray30"),
    axis.title = element_text(face = "bold", size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.text.y = element_text(size = 10, face = "bold"),
    panel.grid = element_blank(),
    legend.position = "none"
  )

print(phm)

# Create summary bar plots with official colors
p1 <- ggplot(assay_counts, aes(x = reorder(assay, n_tissues), y = n_tissues, fill = assay)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = n_tissues), hjust = -0.3, fontface = "bold", size = 3.5) +
  coord_flip() +
  scale_fill_manual(values = assay_colors_used) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Datasets by Assay", x = NULL, y = "Number of Tissues") +
  theme_classic() +
  theme(
    plot.title = element_text(face = "bold", size = 12, hjust = 0.5),
    axis.text.y = element_text(face = "bold", size = 9),
    legend.position = "none"
  )

p2 <- ggplot(tissue_counts, aes(x = reorder(tissue, n_assays), y = n_assays, fill = tissue)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = n_assays), hjust = -0.3, fontface = "bold", size = 3) +
  coord_flip() +
  scale_fill_manual(values = tissue_colors_used) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Datasets by Tissue", x = NULL, y = "Number of Assays") +
  theme_classic() +
  theme(
    plot.title = element_text(face = "bold", size = 12, hjust = 0.5),
    axis.text.y = element_text(face = "bold", size = 8),
    legend.position = "none"
  )

# Print summary information
cat("\n=== Dataset Summary ===\n")
cat("Total unique assay-tissue combinations:", nrow(dataset_summary), "\n")
cat("Number of assays:", length(unique(dataset_summary$assay)), "\n")
cat("Number of tissues:", length(unique(dataset_summary$tissue)), "\n\n")

cat("Assays available:\n")
print(sort(unique(dataset_summary$assay)))

cat("\nTissues available:\n")
print(sort(unique(dataset_summary$tissue)))

# Display side-by-side bar plots
gridExtra::grid.arrange(p1, p2, ncol = 2)
```

## Map Gene Symbols to Features

Now we need to map our gene symbols to the various feature identifiers used across different omics platforms:

```{r map-genes}
# Load feature-to-gene mapping
data(FEATURE_TO_GENE)

# Filter for our genes of interest
gene_features <- FEATURE_TO_GENE %>%
  filter(gene_symbol %in% custom_genes) %>%
  select(gene_symbol, feature_ID) %>%
  distinct()

cat("Found", nrow(gene_features), "feature-to-gene mappings for",
    length(unique(gene_features$gene_symbol)), "of your",
    length(custom_genes), "input genes\n\n")
```

## Extract Results for Your Genes

Filter the differential analysis results to include only your genes of interest:

```{r filter-results}
# Join with differential analysis results
custom_gene_results <- da_results %>%
  inner_join(gene_features, by = "feature_ID") %>%
  filter(!is.na(logFC), !is.na(adj_p_value))

cat("Retrieved", nrow(custom_gene_results), "differential analysis results\n")
cat("for", length(unique(custom_gene_results$gene_symbol)), "genes\n")
cat("across", length(unique(custom_gene_results$tissue)), "tissues\n")
cat("and", length(unique(custom_gene_results$assay)), "omics platforms\n\n")

# Show gene coverage across omics platforms
gene_omics_summary <- custom_gene_results %>%
  select(gene_symbol, assay) %>%
  distinct() %>%
  group_by(gene_symbol) %>%
  summarise(omics_layers = paste(sort(unique(assay)), collapse = ", "),
            n_omics = n(),
            .groups = "drop") %>%
  arrange(desc(n_omics), gene_symbol)

cat("Gene coverage across omics platforms:\n")
print(gene_omics_summary)
cat("\n")

# Check which genes have significant results (adj_p < 0.05)
sig_genes <- custom_gene_results %>%
  filter(adj_p_value < 0.05) %>%
  group_by(gene_symbol) %>%
  summarise(
    n_significant = n(),
    n_tissues = n_distinct(tissue),
    n_omics = n_distinct(assay),
    .groups = "drop"
  ) %>%
  arrange(desc(n_significant))

cat("Genes with significant training effects (FDR < 0.05):\n")
print(sig_genes)
```

# Visualization 1: Summary Tables

## Create a comprehensive table of results

Display logFC and adjusted p-values for each gene across tissues and time points.

**How to interpret this table:**

- **logFC (log2 Fold Change)**: Positive values = upregulation with training; negative values = downregulation
- **Adjusted p-value**: FDR-corrected p-values; values < 0.05 are typically considered significant
- **Comparison groups**: 1w, 2w, 4w, 8w represent weeks of endurance training
- **Use the filters** at the top of each column to search for specific genes, tissues, or omics
- **Sort** by clicking column headers to find the largest changes or most significant results

```{r summary-table}
# Create a detailed summary table
summary_table <- custom_gene_results %>%
  select(gene_symbol, assay, tissue, sex, comparison_group, logFC, adj_p_value) %>%
  mutate(
    logFC = round(logFC, 3),
    adj_p_value = round(adj_p_value, 4),
    significance = case_when(
      adj_p_value < 0.001 ~ "***",
      adj_p_value < 0.01 ~ "**",
      adj_p_value < 0.05 ~ "*",
      TRUE ~ ""
    )
  ) %>%
  arrange(gene_symbol, assay, tissue, sex, comparison_group)

# Display interactive table
DT::datatable(
  summary_table,
  filter = 'top',
  options = list(
    pageLength = 25,
    scrollX = TRUE,
    autoWidth = TRUE
  ),
  caption = "Differential Analysis Results for Custom Gene List. Significance: *** p<0.001, ** p<0.01, * p<0.05"
) %>%
  DT::formatStyle(
    'adj_p_value',
    backgroundColor = DT::styleInterval(c(0.05), c('#ffcccc', '#ffffff'))
  )
```



# Visualization 2: Temporal Trajectories

## Line Plots for Individual Genes

Use `plot_feature_logfc()` to visualize how each gene's expression changes over the training time course:

```{r line-plots, fig.width=10, fig.height=6}
# Get unique features for plotting
plot_features <- custom_gene_results %>%
  select(gene_symbol, assay, tissue, feature_ID) %>%
  distinct() %>%
  # Focus on transcriptomics and proteomics for clearer visualization
  filter(assay %in% c("TRNSCRPT"))

# Plot temporal trajectories for each gene-tissue-omics combination
for (i in 1:min(12, nrow(plot_features))) {  # Limit to first 12 for demonstration
  gene <- plot_features$gene_symbol[i]
  tissue <- plot_features$tissue[i]
  assay <- plot_features$assay[i]
  feature_id <- plot_features$feature_ID[i]

  cat("\nPlotting:", gene, "-", assay, "-", tissue, "\n")

  tryCatch({
    p <- plot_feature_logfc(
      assay = assay,
      tissue = tissue,
      feature_ID = feature_id,
      add_gene_symbol = TRUE,
      facet_by_sex = FALSE
    )
    print(p)
  }, error = function(e) {
    cat("Could not plot:", gene, "-", assay, "-", tissue, "\n")
  })
}
```

## Multi-Gene Trajectory Comparison

Compare temporal patterns across all tissues for your gene of interest.

**How to interpret this plot:**

- **Line trajectories** show how gene expression changes over the training time course (1w to 8w)
- **Point size** indicates significance: larger points = more significant (-log10 of FDR)
- **Colors** represent different tissues (using official MoTrPAC tissue colors)
- **Background shading**: Gray area = downregulation, white area = upregulation
- **Compare panels**: Female (left) vs. Male (right) to identify sex-specific responses

**Look for these patterns:**

- **Sustained response**: Consistent up or down across all time points
- **Early transient**: Change at 1-2 weeks that returns to baseline
- **Late response**: Change only at 4-8 weeks (chronic adaptation)
- **Biphasic**: Initial change in one direction, then reversal
- **Tissue specificity**: Strong response in metabolic tissues (muscle, liver) vs. others

```{r multi-gene-trajectory, fig.width=14, fig.height=7}
# Use the first gene from the custom gene list
selected_gene <- custom_genes[1]

trajectory_data <- custom_gene_results %>%
  filter(
    gene_symbol == selected_gene,
    assay == "TRNSCRPT",
    !is.na(comparison_group)
  )

if (nrow(trajectory_data) > 0) {
  # Convert comparison_group to factor for proper ordering
  trajectory_data <- trajectory_data %>%
    mutate(
      comparison_group = factor(comparison_group, levels = c("1w", "2w", "4w", "8w")),
      tissue = factor(tissue),
      sex = factor(sex, levels = c("female", "male"))
    )

  # Get tissue colors for available tissues
  tissues_in_data <- unique(trajectory_data$tissue)
  tissue_colors_plot <- TISSUE_COLORS[names(TISSUE_COLORS) %in% tissues_in_data]

  # Create a beautiful publication-ready plot
  p <- ggplot(trajectory_data, aes(x = comparison_group, y = logFC,
                                    color = tissue, group = tissue)) +
    # Add subtle background
    geom_rect(
      aes(xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = 0),
      fill = "gray95", alpha = 0.5, inherit.aes = FALSE
    ) +
    geom_rect(
      aes(xmin = -Inf, xmax = Inf, ymin = 0, ymax = Inf),
      fill = "white", alpha = 0.5, inherit.aes = FALSE
    ) +
    # Reference line at zero
    geom_hline(yintercept = 0, linetype = "solid", color = "gray40", linewidth = 0.8) +
    # Lines with shadow effect
    geom_line(linewidth = 2.5, alpha = 0.2) +
    geom_line(linewidth = 1.5, alpha = 0.9) +
    # Points with outline
    geom_point(aes(size = -log10(adj_p_value)),
               alpha = 0.95, shape = 21, fill = "white", stroke = 1.5) +
    geom_point(aes(size = -log10(adj_p_value)),
               alpha = 0.85, shape = 16) +
    # Faceting
    facet_wrap(~ sex, labeller = labeller(sex = c(female = "Female", male = "Male"))) +
    # Colors
    scale_color_manual(values = tissue_colors_plot) +
    scale_size_continuous(
      range = c(3, 10),
      breaks = c(1, 2, 3, 5),
      labels = c("1", "2", "3", "5+")
    ) +
    # Labels
    labs(
      title = bquote(bold(italic(.(selected_gene))) ~ "Expression Across Tissues"),
      subtitle = "Transcriptomics (RNA-seq) temporal response to endurance training",
      x = "Training Duration",
      y = expression(bold(log[2] ~ "Fold Change")),
      color = "Tissue",
      size = expression(-log[10] ~ "(FDR)")
    ) +
    # Theme
    theme_minimal(base_size = 13) +
    theme(
      # Plot titles
      plot.title = element_text(face = "bold", size = 18, hjust = 0.5, margin = margin(b = 5)),
      plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray40",
                                   margin = margin(b = 15), face = "italic"),
      plot.background = element_rect(fill = "white", color = NA),

      # Axes
      axis.title.x = element_text(face = "bold", size = 13, margin = margin(t = 10)),
      axis.title.y = element_text(face = "bold", size = 13, margin = margin(r = 10)),
      axis.text = element_text(size = 11, color = "black"),
      axis.text.x = element_text(face = "bold"),
      axis.line = element_line(color = "gray30", linewidth = 0.5),

      # Facet strips
      strip.background = element_rect(fill = "gray20", color = "gray20", linewidth = 0),
      strip.text = element_text(face = "bold", size = 13, color = "white",
                               margin = margin(5, 5, 5, 5)),

      # Panel
      panel.background = element_rect(fill = "white", color = NA),
      panel.border = element_rect(color = "gray30", fill = NA, linewidth = 1),
      panel.grid.major.y = element_line(color = "gray85", linewidth = 0.4, linetype = "dotted"),
      panel.grid.major.x = element_blank(),
      panel.grid.minor = element_blank(),
      panel.spacing = unit(1.5, "lines"),

      # Legend
      legend.position = "right",
      legend.background = element_rect(fill = "gray98", color = "gray60", linewidth = 0.5),
      legend.title = element_text(face = "bold", size = 12),
      legend.text = element_text(size = 10),
      legend.key = element_rect(fill = "white", color = NA),
      legend.spacing.y = unit(0.3, "cm"),
      legend.margin = margin(10, 10, 10, 10)
    ) +
    # Guides
    guides(
      color = guide_legend(override.aes = list(size = 5, alpha = 1),
                          keyheight = unit(0.8, "cm")),
      size = guide_legend(keyheight = unit(0.6, "cm"))
    )

  print(p)
} else {
  cat("No data available for", selected_gene, "in TRNSCRPT assay\n")
}
```

# Visualization 3: Heatmaps

## Heatmap of logFC Across Tissues and Time Points

Create heatmaps showing the pattern of regulation for each gene across tissues, with female and male data side-by-side.

**How to interpret these heatmaps:**

- **Color intensity** represents magnitude and direction of change:
  - **Blue** = downregulation (negative logFC)
  - **Red** = upregulation (positive logFC)
  - **White** = no change
- **Asterisks (\*)** mark statistically significant changes (FDR < 0.05)
- **Column arrangement**: Female time points (1w-8w) | Male time points (1w-8w)
- **Rows are NOT clustered** to preserve tissue order for easier comparison between genes

**What to look for:**

- **Horizontal patterns** (across time): How does the response evolve over training?
- **Vertical patterns** (across tissues): Which tissues show similar responses?
- **Sex differences**: Compare left half (female) vs. right half (male)
- **Consistency**: Are changes significant (*) across multiple time points?

```{r heatmap-tissue-time, fig.height=8}
# Function to create heatmap for a specific gene and omics layer (both sexes)
create_gene_heatmap <- function(gene_name, omics_layer) {

  # Get both logFC and p-values
  heatmap_full_data <- custom_gene_results %>%
    filter(
      gene_symbol == gene_name,
      assay == omics_layer,
      !is.na(comparison_group)
    ) %>%
    select(tissue, comparison_group, sex, logFC, adj_p_value) %>%
    # Handle duplicates by taking the mean
    group_by(tissue, comparison_group, sex) %>%
    summarise(
      logFC = mean(logFC, na.rm = TRUE),
      adj_p_value = mean(adj_p_value, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    # Create combined column for time point and sex
    mutate(timepoint_sex = paste0(comparison_group, "_", sex))

  # Create logFC matrix
  heatmap_data <- heatmap_full_data %>%
    select(tissue, timepoint_sex, logFC) %>%
    pivot_wider(
      names_from = timepoint_sex,
      values_from = logFC,
      values_fill = 0
    ) %>%
    column_to_rownames("tissue")

  # Create significance matrix
  sig_data <- heatmap_full_data %>%
    select(tissue, timepoint_sex, adj_p_value) %>%
    pivot_wider(
      names_from = timepoint_sex,
      values_from = adj_p_value,
      values_fill = 1
    ) %>%
    column_to_rownames("tissue")

  if (nrow(heatmap_data) == 0) {
    cat("No data available for", gene_name, "-", omics_layer, "\n")
    return(NULL)
  }

  # Order columns: female time points, then male time points
  time_order <- c("1w", "2w", "4w", "8w")
  female_cols <- paste0(time_order, "_female")
  male_cols <- paste0(time_order, "_male")
  desired_order <- c(female_cols, male_cols)

  # Keep only columns that exist in the data
  available_cols <- intersect(desired_order, colnames(heatmap_data))
  heatmap_data <- heatmap_data[, available_cols, drop = FALSE]
  sig_data <- sig_data[, available_cols, drop = FALSE]

  # Create display labels matrix with asterisks for significant values
  display_labels <- matrix("", nrow = nrow(heatmap_data), ncol = ncol(heatmap_data))
  display_labels[sig_data < 0.05] <- "*"
  rownames(display_labels) <- rownames(heatmap_data)
  colnames(display_labels) <- colnames(heatmap_data)

  # Get max absolute value for color scale, ensure it's not zero
  max_abs_val <- max(abs(heatmap_data), na.rm = TRUE)
  if (max_abs_val == 0) max_abs_val <- 1

  # Create annotation for column groups
  col_annotation <- data.frame(
    Sex = ifelse(grepl("_female$", colnames(heatmap_data)), "Female", "Male"),
    row.names = colnames(heatmap_data)
  )

  ann_colors <- list(Sex = c(Female = "#FF69B4", Male = "#4169E1"))

  # Create heatmap without clustering, with asterisks for significant values
  print(pheatmap(
    heatmap_data,
    cluster_rows = FALSE,
    cluster_cols = FALSE,
    scale = "none",
    color = colorRampPalette(c("blue", "white", "red"))(100),
    breaks = seq(-max_abs_val, max_abs_val, length.out = 101),
    annotation_col = col_annotation,
    annotation_colors = ann_colors,
    display_numbers = display_labels,
    number_color = "black",
    fontsize_number = 14,
    main = paste0(gene_name, " - ", omics_layer, "\nlogFC Across Tissues and Time (* FDR < 0.05)"),
    fontsize = 10,
    fontsize_row = 9,
    fontsize_col = 9,
    cellwidth = 25,
    cellheight = 18,
    border_color = "gray80",
    main_fontsize = 11
  ))
}

# Create heatmaps for the first three genes in the list
genes_to_plot <- custom_genes[1:min(3, length(custom_genes))]

for (gene in genes_to_plot) {
  if (gene %in% custom_gene_results$gene_symbol) {
    # Transcriptomics
    if ("TRNSCRPT" %in% (custom_gene_results %>% filter(gene_symbol == gene) %>% pull(assay))) {
      create_gene_heatmap(gene, "TRNSCRPT")
    }

    # Proteomics
    if ("PROT" %in% (custom_gene_results %>% filter(gene_symbol == gene) %>% pull(assay))) {
      create_gene_heatmap(gene, "PROT")
    }
  }
}
```

## Combined Heatmap: All Genes, Selected Tissue

Compare all genes simultaneously in a specific tissue, showing both female and male responses side-by-side.

**How to interpret this combined heatmap:**

- **Each row** represents one gene from your custom list
- **Color patterns** show regulation across the time course:
  - **Blue** = downregulation
  - **Red** = upregulation
  - **Gray** = no data available for that combination
- **Asterisks (\*)** mark significant changes (FDR < 0.05)
- **Column organization**: Female time points (1w-8w) | Male time points (1w-8w)

**What to look for:**

- **Co-regulation**: Do multiple genes show similar patterns? (same color patterns across rows)
- **Sex-specific responses**: Are there genes that respond only in females or only in males?
- **Time course patterns**: Do genes show early (1-2w) or late (4-8w) responses?
- **Consistency of significance**: Are the strongest color changes also marked with asterisks?

```{r heatmap-all-genes, fig.height=7}
# Select tissue and assay for comparison
selected_tissue <- "HEART"
selected_assay <- "TRNSCRPT"

# Get both logFC and p-values
combined_full_data <- custom_gene_results %>%
  filter(
    tissue == selected_tissue,
    assay == selected_assay,
    !is.na(comparison_group)
  ) %>%
  select(gene_symbol, comparison_group, sex, logFC, adj_p_value) %>%
  # Handle duplicates by taking the mean
  group_by(gene_symbol, comparison_group, sex) %>%
  summarise(
    logFC = mean(logFC, na.rm = TRUE),
    adj_p_value = mean(adj_p_value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  # Create combined column for time point and sex
  mutate(timepoint_sex = paste0(comparison_group, "_", sex))

# Create logFC matrix
combined_data <- combined_full_data %>%
  select(gene_symbol, timepoint_sex, logFC) %>%
  pivot_wider(
    names_from = timepoint_sex,
    values_from = logFC,
    values_fill = NA
  ) %>%
  column_to_rownames("gene_symbol")

# Create significance matrix
combined_sig_data <- combined_full_data %>%
  select(gene_symbol, timepoint_sex, adj_p_value) %>%
  pivot_wider(
    names_from = timepoint_sex,
    values_from = adj_p_value,
    values_fill = 1
  ) %>%
  column_to_rownames("gene_symbol")

if (nrow(combined_data) > 0) {
  # Order columns: female time points, then male time points
  time_order <- c("1w", "2w", "4w", "8w")
  female_cols <- paste0(time_order, "_female")
  male_cols <- paste0(time_order, "_male")
  desired_order <- c(female_cols, male_cols)

  # Keep only columns that exist in the data
  available_cols <- intersect(desired_order, colnames(combined_data))
  combined_data <- combined_data[, available_cols, drop = FALSE]
  combined_sig_data <- combined_sig_data[, available_cols, drop = FALSE]

  # Create display labels matrix with asterisks for significant values
  combined_display_labels <- matrix("", nrow = nrow(combined_data), ncol = ncol(combined_data))
  combined_display_labels[combined_sig_data < 0.05] <- "*"
  rownames(combined_display_labels) <- rownames(combined_data)
  colnames(combined_display_labels) <- colnames(combined_data)

  # Create annotation for column groups
  col_annotation <- data.frame(
    Sex = ifelse(grepl("_female$", colnames(combined_data)), "Female", "Male"),
    row.names = colnames(combined_data)
  )

  ann_colors <- list(Sex = c(Female = "#FF69B4", Male = "#4169E1"))

  pheatmap(
    combined_data,
    cluster_rows = FALSE,
    cluster_cols = FALSE,
    scale = "none",
    color = colorRampPalette(c("blue", "white", "red"))(100),
    breaks = seq(-3, 3, length.out = 101),
    annotation_col = col_annotation,
    annotation_colors = ann_colors,
    display_numbers = combined_display_labels,
    number_color = "black",
    fontsize_number = 14,
    main = paste0("All Genes in ", selected_tissue, " (", selected_assay, ")\nlogFC Across Training Time Course (* FDR < 0.05)"),
    fontsize = 10,
    fontsize_row = 9,
    fontsize_col = 9,
    cellwidth = 30,
    cellheight = 15,
    na_col = "grey90",
    border_color = "gray80",
    main_fontsize = 11
  )
} else {
  cat("No data available for the selected combination\n")
}
```

# Next Steps

## Customizing This Workflow

You can adapt this vignette by:

1. **Changing the gene list**: Replace `custom_genes` with your genes of interest
2. **Focusing on specific tissues**: Modify `selected_tissue` in the heatmap sections
3. **Selecting omics layers**: Focus on specific assays (TRNSCRPT, PROT, PHOSPHO, ACETYL, UBIQ, METAB)
4. **Adjusting significance thresholds**: Change the 0.05 FDR cutoff
5. **Adding normalized data**: Use `combine_normalized_data()` to get actual expression values

## Getting Help

For questions about the MoTrPAC packages:

- [MotrpacRatTraining6moData issues](https://github.com/MoTrPAC/MotrpacRatTraining6moData/issues)
- [MotrpacRatTraining6mo issues](https://github.com/MoTrPAC/MotrpacRatTraining6mo/issues)
- [Workshop: MoTrPAC R Packages from the Endurance Training in Young Rats Study](https://motrpac.github.io/MotrpacWorkshops/docs/articles/rat-endurance-6m.html)


